# Final Project - Part 2


Datasets: 

https://www.kaggle.com/datasets/econdata/climate-change/data
https://www.e-education.psu.edu/meteo815/node/525
https://ourworldindata.org/co2-dataset-sources

```{R}
climate_data = read.csv(("C:/Users/mgrod/Documents/R_Data_Science/Final_Project_Stats/Final_Project/climate_change_yearly_averages.csv"), sep = ",")

summary(climate_data)
```
```{r}
#plot each variable against sea level to inspect individual relationships


#MEI vs. Sea.Level
plot(climate_data$MEI,climate_data$Sea.Level,xlab="MEI",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(-1,2))

#CO2 vs. Sea.Level
plot(climate_data$CO2,climate_data$Sea.Level,xlab="CO2",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(340,382))

#CH4 vs. Sea.Level
plot(climate_data$CH4,climate_data$Sea.Level,xlab="CH4",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(1640,1800))

#N2O vs. Sea.Level
plot(climate_data$N2O,climate_data$Sea.Level,xlab="N2O",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(300,320))

#CFC.11 vs. Sea.Level
plot(climate_data$CFC.11,climate_data$Sea.Level,xlab="CFC.11",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(193,272))

#CFC.12 vs. Sea.Level
plot(climate_data$CFC.12,climate_data$Sea.Level,xlab="CFC.12",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(356,544))

#TSI vs. Sea.Level
plot(climate_data$TSI,climate_data$Sea.Level,xlab="TSI",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(1365,1367))

#Aerosols vs. Sea.Level
plot(climate_data$Aerosols,climate_data$Sea.Level,xlab="Aerosols",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(0,0.15))

#Temp vs. Sea.Level
plot(climate_data$Temp,climate_data$Sea.Level,xlab="Temp",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(-0.1,0.6))

#CO2 Emissions vs. Sea.Level
plot(climate_data$CO2.Emissions,climate_data$Sea.Level,xlab="CO2 Emissions",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(18,31))
```


```{R}

#regression fit
fit <- lm(Sea.Level ~ MEI+CO2+CH4+N2O+CFC.11+CFC.12+TSI+Aerosols+Temp+CO2.Emissions, data=climate_data)
summary(fit)



#Look for violations of Constant Variance
#########################################

#plot all residuals
plot (fitted (fit), residuals (fit), xlab="Fitted", ylab="Residuals")
abline (h=0)
plot (fitted (fit), abs (residuals (fit)),
xlab="Fitted", ylab="|Residuals|")

#plot residuals against individual predictors on one plot side-by-side and absolute values on four individual plots
car::residualPlots(fit, pch=20, col="red", fitted = F, ask = F, layout = c(3,2), tests = F, quadratic = F)
plot (climate_data$MEI, abs(residuals (fit)), xlab="MEI", ylab="}|Residuals|")
plot (climate_data$CO2, abs(residuals (fit)), xlab="CO2", ylab="}|Residuals|")
plot (climate_data$CH4, abs(residuals (fit)), xlab="CH4", ylab="}|Residuals|")
plot (climate_data$N2O, abs(residuals (fit)), xlab="N2O", ylab="}|Residuals|")
plot (climate_data$CFC.11, abs(residuals (fit)), xlab="CFC.11", ylab="}|Residuals|")
plot (climate_data$CFC.12, abs(residuals (fit)), xlab="CFC.12", ylab="}|Residuals|")
plot (climate_data$TSI, abs(residuals (fit)), xlab="TSI", ylab="}|Residuals|")
plot (climate_data$Aerosols, abs(residuals (fit)), xlab="Aerosols", ylab="}|Residuals|")
plot (climate_data$Temp, abs(residuals (fit)), xlab="Temp", ylab="}|Residuals|")
plot (climate_data$CO2.Emissions, abs(residuals (fit)), xlab="CO2.Emissions", ylab="}|Residuals|")

#two different populations can be seen for CF.12 below and above 500, let's test them
var.test (residuals(fit)[climate_data$CFC.12>500], residuals(fit)[climate_data$CFC.12<=500])
#no significant difference detected by variance test indicating constant variance



#Look for violations of Normality
#########################################
res <- fit$residuals
#histogram of residuals
hist(res, xlab = "Residuals", probability = T, ylim = c(0, 6))
#empirical density
lines(density(res, na.rm=T), lwd = 2, col = "red")
#best fitting normal curve
curve(dnorm(x, mean = mean(res, na.rm=T), sd = sd(res, na.rm=T)), lty = 2, lwd = 2, add = TRUE, col = "blue")
#small deviation from Normality seen with longer lower tail 

#ordered values of the residuals vs quantiles of the standard normal distribution (Q-Q plot)
qqnorm(as.numeric(res), col="red", pch=20)
qqline(as.numeric(res), col="blue", lty=1, lwd=1)
#deviation from standard normal distribution seen in the bottom quantile

#The Shapiro-Wilk test 
shapiro.test (residuals (fit))
#no significant violation of Normality detected


#Look for violations of Linearity
#########################################
#generate component-plus-residual plot for each predictor
car::crPlots(fit, terms = . ~ MEI+CO2+CH4+N2O+CFC.11+CFC.12+TSI+Aerosols+Temp+CO2.Emissions, pch=20, col="red")
#small deviation from linearity seen for TSI and Aerosols


#Look for Correlations
#########################################
#make an index plot of the residuals. Look for longer runs of residuals above or below the line.
plot (residuals (fit), ylab="Residuals")
abline (h=0)
#no significant correlation seen

#check for correlation using a regression of successive residuals
summary(lm(residuals(fit)[-2] ~ 1+residuals (fit) [-1]))
#significant correlation seen

#compute the Durbin-Watson statistic
library(lmtest)
dwtest (Sea.Level ~ MEI+CO2+CH4+N2O+CFC.11+CFC.12+TSI+Aerosols+Temp+CO2.Emissions, data=climate_data)
#no significant correlation seen


#Check measurement validity 
#########################################

#Inspect histograms
hist(climate_data$MEI, main = "", xlab = "MEI")
hist(climate_data$CO2, main = "", xlab = "CO2")
hist(climate_data$CH4, main = "", xlab = "CH4")
hist(climate_data$N2O, main = "", xlab = "N2O")
hist(climate_data$CFC.11, main = "", xlab = "CFC.11")
hist(climate_data$CFC.12, main = "", xlab = "CFC.12")
hist(climate_data$TSI, main = "", xlab = "TSI")
hist(climate_data$Aerosols, main = "", xlab = "Aerosols")
hist(climate_data$Temp, main = "", xlab = "Temp")
hist(climate_data$CO2.Emissions, main = "", xlab = "CO2.Emissions")


```






```{R}

#Look for outliers
#########################################
#Compute Studentized residuals
resstudent <- rstudent(fit)
#Plot Studentized residuals
plot (resstudent, ylab="Studentized Residuals")
abline (h=0)
#test Studentized residuals (no observations indicated to be an outlier)
car::outlierTest(fit, n.max = Inf)


#Look for potentially influential points
#########################################
#compute and inspect Cook's statistics
cook <- cooks.distance (fit)
cook
library(ggplot2)
library(faraway)
halfnorm (cook, 3, ylab="Cook's distances")

#point 5 with the largest Cook's distances is indicated as the most influential

#exclude point 5 with the largest Cook's distance and therefore the most influential and re-fit the data
fit_minus_5 <- lm (Sea.Level ~ MEI+CO2+CH4+N2O+CFC.11+CFC.12+TSI+Aerosols+Temp+CO2.Emissions, climate_data, subset=(cook < max (cook)))
summary (fit_minus_5)

#compare with fitting the full dataset
fit_full <- lm (Sea.Level ~ MEI+CO2+CH4+N2O+CFC.11+CFC.12+TSI+Aerosols+Temp+CO2.Emissions, data=climate_data)
summary (fit_full)

#the model with point 5 removed produces slightly better fit than the full model as evidenced by:
# 1. higher R-squared value
# 2. higher F-statistic


#Look for leverage points
#########################################
#compute the leverage value for all the observations
leverage = hatvalues(fit)

#plot leverage values
plot(leverage)

library(car)

#plot residuals, leverage, and Cook's distance 
influencePlot(fit)

# point 5 is being identified as high leverage point


#Look again for potentially influential points
#########################################
#compute and inspect Cook's statistics
cook <- cooks.distance (fit_minus_5)
halfnorm (cook, 3, ylab="Cook's distances")



#Look again for leverage points
#########################################
#compute the leverage value for all the observations
leverage = hatvalues(fit_minus_5)

#plot leverage values
plot(leverage)

#plot residuals, leverage, and Cook's distance 
influencePlot(fit_minus_5)
```

An influential observation is an observation that if included in the fit will modify the estimated coefficients of the regression model by a significant amount. This can be quantified by the Cook’s distance which measures how much the estimated coefficients change when an observation is removed from the regression model.

A high leverage point is an observation with a value of the independent variables far from the line of best fit. A point could be either influential or high leverage or both.
 
An outlier is an observation with a large residual relative to its standard deviation. An outlier usually signifies a measurement error.


```{r}
#create test and training datasets
library(MASS)
data(climate_data)

set.seed(11)
n = floor(0.8 * nrow(climate_data))
index = sample(seq_len(nrow(climate_data)), size = n)

train = climate_data[index, ]
test = climate_data[-index, ]
head(train)
summary(train)
head(test)
summary(test)

```
```{r}
# fit a model to the training dataset
lm_train <- lm(Sea.Level ~ MEI+CO2+CH4+N2O+CFC.11+CFC.12+TSI+Aerosols+Temp+CO2.Emissions, data=train)
summary(lm_train)

par(mfrow = c(2,2))
plot(fit)
df = data.frame(x = fitted(lm_train), y = stdres(lm_train))
ggplot(df, aes(x = x, y = y)) + 
    geom_point() + 
    theme_bw() + 
    geom_hline(yintercept = 0)

#MSPE with test data
MPSE1 = mean((test$Temp - predict.lm(lm_train, test))^2)
MPSE1

```
```{r}
#remove MEI

lm_train2 <- update(lm_train, .~. - MEI)
summary(lm_train2)

par(mfrow = c(2,2))
plot(lm_train2)
df2 = data.frame(x = fitted(lm_train2), y = stdres(lm_train2))
ggplot(df2, aes(x = x, y = y)) + 
    geom_point() + 
    theme_bw() + 
    geom_hline(yintercept = 0)

#MSPE with test data
MPSE2 = mean((test$Temp - predict.lm(lm_train2, test))^2)
MPSE2

#This second linear model (lm_train2) is better as it has higher F-statistic, and R-squared and smaller MSPE than lm_train

```
```{r}
#remove CH4

lm_train3 <- update(lm_train2, .~. - CH4)
summary(lm_train3)

par(mfrow = c(2,2))
plot(lm_train3)
df3 = data.frame(x = fitted(lm_train3), y = stdres(lm_train3))
ggplot(df3, aes(x = x, y = y)) + 
    geom_point() + 
    theme_bw() + 
    geom_hline(yintercept = 0)

#MSPE with test data
MPSE3 = mean((test$Temp - predict.lm(lm_train3, test))^2)
MPSE3

#This third linear model (lm_train3) is better as it has higher F-statistic, and R-squared than lm_train2. Also more coefficients are significant. However, MPSE is higher.
```
```{r}
#remove Temp

lm_train4 <- update(lm_train3, .~. - Temp)
summary(lm_train4)

par(mfrow = c(2,2))
plot(lm_train4)
df4 = data.frame(x = fitted(lm_train4), y = stdres(lm_train4))
ggplot(df4, aes(x = x, y = y)) + 
    geom_point() + 
    theme_bw() + 
    geom_hline(yintercept = 0)

#MSPE with test data
MPSE4 = mean((test$Temp - predict.lm(lm_train4, test))^2)
MPSE4

#This fourth linear model (lm_train4) is better as it has higher F-statistic, and R-squared than lm_train3. Also coefficients are more significant.  However, MPSE is higher.

```


```{r}
#calculate AIC and BIC for each of the three models
n = dim(test)[1]; 

#LM (lm_train)
rss <- sum(resid(lm_train)^2)
AIC_lm_train = 2*6 + n*log(rss/n)
BIC_lm_train = log(n)*6 + n*log(rss/n)
AIC_lm_train
BIC_lm_train

#LM (lm_train2)
rss <- sum(resid(lm_train2)^2)
AIC_lm_train2 = 2*6 + n*log(rss/n)
BIC_lm_train2 = log(n)*6 + n*log(rss/n)
AIC_lm_train2
BIC_lm_train2

#LM (lm_train3)
rss <- sum(resid(lm_train3)^2)
AIC_lm_train3 = 2*6 + n*log(rss/n)
BIC_lm_train3 = log(n)*6 + n*log(rss/n)
AIC_lm_train3
BIC_lm_train3


#LM (lm_train4)
rss <- sum(resid(lm_train4)^2)
AIC_lm_train4 = 2*6 + n*log(rss/n)
BIC_lm_train4 = log(n)*6 + n*log(rss/n)
AIC_lm_train4
BIC_lm_train4



#Both BIC and AIC indicate that the model with all variables (lm_train) is the best choice
```
```{r}
require(MASS)
VIFs = vif(lm_train)
VIFs
K = kappa(lm_train)
K
cor(model.matrix(lm_train)[,-1])
```
```{r}
col4 = colorRampPalette(c("black", "darkgrey", "grey","#CFB87C"))
#corrplot(cor(climate_data[2:12]), method = "ellipse", col = col4(12),  addCoef.col = "black", tl.col = "black", number.cex=0.75)


```
```{r}
#CO2 Emissions and CO2 are highly correlated at 0.98. We should retain only one of the two going forward.
#CO2 has the largest VIF. Let's remove it.
lm_train5 <- update(lm_train, .~. - CO2)
summary(lm_train5)

par(mfrow = c(2,2))
plot(lm_train5)
df5 = data.frame(x = fitted(lm_train4), y = stdres(lm_train5))
ggplot(df5, aes(x = x, y = y)) + 
    geom_point() + 
    theme_bw() + 
    geom_hline(yintercept = 0)

#MSPE with test data
MPSE5 = mean((test$Temp - predict.lm(lm_train5, test))^2)
MPSE5
```
```{r}
VIFs = vif(lm_train5)
VIFs
K = kappa(lm_train5)
K
cor(model.matrix(lm_train5)[,-1])

```
```{r}

#CFC.12 has the largest VIF. Let's remove it.
lm_train6 <- update(lm_train5, .~. - CFC.12)
summary(lm_train6)

par(mfrow = c(2,2))
plot(lm_train6)
df6 = data.frame(x = fitted(lm_train6), y = stdres(lm_train6))
ggplot(df6, aes(x = x, y = y)) + 
    geom_point() + 
    theme_bw() + 
    geom_hline(yintercept = 0)

#MSPE with test data
MPSE6 = mean((test$Temp - predict.lm(lm_train6, test))^2)
MPSE6

#Let's check for additional collinearities
VIFs = vif(lm_train6)
VIFs
K = kappa(lm_train6)
K
cor(model.matrix(lm_train6)[,-1])

```
```{r}

#CH4 has the largest VIF. Let's remove it.
lm_train7 <- update(lm_train6, .~. - CH4)
summary(lm_train7)

par(mfrow = c(2,2))
plot(lm_train7)
df7 = data.frame(x = fitted(lm_train7), y = stdres(lm_train7))
ggplot(df7, aes(x = x, y = y)) + 
    geom_point() + 
    theme_bw() + 
    geom_hline(yintercept = 0)

#MSPE with test data
MPSE7 = mean((test$Temp - predict.lm(lm_train7, test))^2)
MPSE7

#Let's check for additional collinearities
VIFs = vif(lm_train7)
VIFs
K = kappa(lm_train7)
K
cor(model.matrix(lm_train7)[,-1])
```
```{r}
#CO2 Emissions, N2O and Temp are all highly correlated. 
#CO2 Emission has the highest VIF. Let's remove it going forward.

lm_train8 <- update(lm_train7, .~. - CO2.Emissions)
summary(lm_train8)

par(mfrow = c(2,2))
plot(lm_train8)
df8 = data.frame(x = fitted(lm_train8), y = stdres(lm_train8))
ggplot(df8, aes(x = x, y = y)) + 
    geom_point() + 
    theme_bw() + 
    geom_hline(yintercept = 0)

#MSPE with test data
MPSE8 = mean((test$Temp - predict.lm(lm_train8, test))^2)
MPSE8

#Let's check for additional collinearities
VIFs = vif(lm_train8)
VIFs
K = kappa(lm_train8)
K
cor(model.matrix(lm_train8)[,-1])
```
```{r}
#fit the reduced model to the entire dataset
fit_final <- lm(Sea.Level ~ MEI+N2O+CFC.11+TSI+Aerosols+Temp, data=climate_data)
summary(fit_final)

#we don't see a good fit. Let's try some non-linear transformations next
```

```{r}

#plot each variable against sea level to inspect individual relationships


#MEI vs. Sea.Level
plot(climate_data$MEI,climate_data$Sea.Level,xlab="MEI",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(-1,2))

#CO2 vs. Sea.Level
plot(climate_data$CO2,climate_data$Sea.Level,xlab="CO2",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(340,382))

#CH4 vs. Sea.Level
plot(climate_data$CH4,climate_data$Sea.Level,xlab="CH4",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(1640,1800))

#N2O vs. Sea.Level
plot(climate_data$N2O,climate_data$Sea.Level,xlab="N2O",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(300,320))

#CFC.11 vs. Sea.Level
plot(climate_data$CFC.11,climate_data$Sea.Level,xlab="CFC.11",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(193,272))

#CFC.12 vs. Sea.Level
plot(climate_data$CFC.12,climate_data$Sea.Level,xlab="CFC.12",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(356,544))

#TSI vs. Sea.Level
plot(climate_data$TSI,climate_data$Sea.Level,xlab="TSI",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(1365,1367))

#Aerosols vs. Sea.Level
plot(climate_data$Aerosols,climate_data$Sea.Level,xlab="Aerosols",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(0,0.15))

#Temp vs. Sea.Level
plot(climate_data$Temp,climate_data$Sea.Level,xlab="Temp",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(-0.1,0.6))

#CO2 Emissions vs. Sea.Level
plot(climate_data$CO2.Emissions,climate_data$Sea.Level,xlab="CO2 Emissions",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(18,31))
```
```{r}
# try transformations on non-linear data

climate_data2 = climate_data
summary(climate_data)

#center
climate_data2$CH4 = scale(climate_data2$CH4, scale=F)
climate_data2$CFC.12 = scale(climate_data2$CFC.12, scale=F)
climate_data2$CO2.Emissions = scale(climate_data2$CO2.Emissions, scale=F)
summary(climate_data2)

#CH4^2 vs. Sea.Level (the relationship is now linear)
plot((climate_data2$CH4+50)^2,climate_data2$Sea.Level,xlab="CH4",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(0,10000))

#CFC.12^2 vs. Sea.Level (the relationship is now more linear)
plot((climate_data2$CFC.12+136)^2,climate_data2$Sea.Level,xlab="CFC.12",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(0,40000))

#log(Aerosols) vs. Sea.Level (the relationship is now linear)
plot(log(climate_data2$Aerosols),climate_data2$Sea.Level,xlab="Aerosols",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(log(0.002),log(0.122)))

#cube root: (CO2.Emissions)^1/3 vs. Sea.Level (the relationship is now linear)
plot(sign(climate_data2$CO2.Emissions)*abs(climate_data2$CO2.Emissions)^(1/3),climate_data2$Sea.Level,xlab="CO2.Emissions",pch=20,ylab="Sea Level", ylim=c(5,8),xlim=c(-3,3))


#transform the data
climate_data2$CH4 = (climate_data2$CH4+50)^2
climate_data2$CFC.12 = (climate_data2$CFC.12+136)^2
climate_data2$Aerosols = log(climate_data2$Aerosols)
climate_data2$CO2.Emissions = sign(climate_data2$CO2.Emissions)*abs(climate_data2$CO2.Emissions)^(1/3)
summary(climate_data2)
```
```{r}

#regression fit on the transformed data
fit_transformed <- lm(Sea.Level ~ MEI+CO2+CH4+N2O+CFC.11+CFC.12+TSI+Aerosols+Temp+CO2.Emissions, data=climate_data2)
summary(fit_transformed)

```
```{r}
col4 = colorRampPalette(c("black", "darkgrey", "grey","#CFB87C"))
corrplot(cor(climate_data2[2:12]), method = "ellipse", col = col4(12),  addCoef.col = "black", tl.col = "black", number.cex=0.75)



```
```{r}

#regression fit on the transformed data (minus CO2, which is correlated with CO2.Emissions, and CFC.12 which is highly non-linear)
fit_transformed2 <- lm(Sea.Level ~ MEI+CH4+N2O+CFC.11+TSI+Aerosols+Temp+CO2.Emissions, data=climate_data2)
summary(fit_transformed2)

```
```{r}
#Let's check for collinearities
VIFs = vif(fit_transformed2)
VIFs
K = kappa(fit_transformed2)
K
cor(model.matrix(fit_transformed2)[,-1])

```
```{r}

#Let's also remove N2O, which is correlated with CO2.Emissions, and has large VIF
fit_transformed3 <- lm(Sea.Level ~ MEI+CH4+CFC.11+TSI+Aerosols+Temp+CO2.Emissions, data=climate_data2)
summary(fit_transformed3)
```
```{r}
#Let's check for collinearities
VIFs = vif(fit_transformed3)
VIFs
K = kappa(fit_transformed3)
K
cor(model.matrix(fit_transformed3)[,-1])


```
```{r}

#Let's also remove CH4, which is correlated with CO2.Emissions, and has large VIF
fit_transformed4 <- lm(Sea.Level ~ MEI+CFC.11+TSI+Aerosols+Temp+CO2.Emissions, data=climate_data2)
summary(fit_transformed4)

```
```{r}
#Let's check for collinearities
VIFs = vif(fit_transformed4)
VIFs
K = kappa(fit_transformed4)
K
cor(model.matrix(fit_transformed4)[,-1])


```

```{r}
#Let's also remove TSI, which is not statistically significant
fit_transformed5 <- lm(Sea.Level ~ MEI+CFC.11+Aerosols+Temp+CO2.Emissions, data=climate_data2)
summary(fit_transformed5)



```

```{r}
#Let's check for collinearities
VIFs = vif(fit_transformed5)
VIFs
K = kappa(fit_transformed5)
K
cor(model.matrix(fit_transformed5)[,-1])

#no significant collinearities remain

```

